---
title: "hw04-dplyr-join-cheatsheet"
author: "My Linh Thibodeau"
date: '2017-10-10'
output: github_document
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
knitr::opts_chunk$set(fig.width=12, fig.height=9)
library(knitr)
library(kableExtra)
library(reshape2)
options(knitr.table.format = "html")
```

# Essentials of data joining with dplyr

# Matching objectives and functions

## Considering two data.frame d1 and d2


| Objective                                           | tidyr/dplyr function    | 
|-----------------------------------------------------|-------------------------|
| Append/join data                                    | MUTATING                | 
|-----------------------------------------------------|-------------------------|
|1. Start with d1 and append matching column(s) of d2 | left_join               | 
|2. Start with d2 and append matching column(s) of d1 | right_join              | 
|3. Only keep overlap d1-d2 based on matching columns | inner_join              |
|4. Keep all of d1 and d2                             | full_join               |
|-----------------------------------------------------|-------------------------|
| Filter data                                         | FILTERING               | 
|-----------------------------------------------------|-------------------------|
|5. Keep rows of d1 if there is a match in d2         | semi_join               |
|4. Discard rows of d1 if there is a match in d2      | anti_join               |

```{r}
d1 <- read.table("small_dataframe_example.txt",sep = "\t", header = TRUE)
d1 %>% kable("html") %>% kable_styling()
d2 <- read.table("small_dataframe_example_2.txt",sep = "\t", header = TRUE)
d2 %>% kable("html") %>% kable_styling()
```


