---
title: "hw04-dplyr-join-cheatsheet"
author: "My Linh Thibodeau"
date: '2017-10-10'
output: github_document
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
knitr::opts_chunk$set(fig.width=12, fig.height=9)
library(knitr)
library(kableExtra)
library(reshape2)
options(knitr.table.format = "html")
```

# Essentials of data joining with dplyr - A table is worth a thousand words

# Matching objectives and functions

## Considering two data.frame d1 and d2


| Objective                                           | tidyr/dplyr function    | 
|-----------------------------------------------------|-------------------------|
| **APPEND/JOIN DATA**                                | **MUTATING**            |
|1. Start with d1 and append matching column(s) of d2 | left_join               | 
|2. Start with d2 and append matching column(s) of d1 | right_join              | 
|3. Keep overlap d1-d2 based on matching column(s)    | inner_join              |
|4. Keep all of d1 and d2                             | full_join               |
| **FILTER DATA**                                     | **FILTERING**           | 
|5. Keep rows of d1 if there is a match in d2         | semi_join               |
|4. Discard rows of d1 if there is a match in d2      | anti_join               |

### DATASET d1

```{r}
d1 <- read.table("scratch-space/small_dataframe_example.txt",sep = "\t", header = TRUE)
d1 %>% kable("html") %>% kable_styling()
```

### DATASET d2

```{r}
d2 <- read.table("scratch-space/small_dataframe_example_2.txt",sep = "\t", header = TRUE)
d2 %>% kable("html") %>% kable_styling()
```
***

## MUTATING JOIN

### (1) left_join

```{r}
left_join(d1, d2, by = "gene") %>% kable("html") %>% kable_styling()
```


### (2) right_join  

```{r}
right_join(d1, d2, by = "gene") %>% kable("html") %>% kable_styling()
```

### (3) inner_join  

```{r}
inner_join(d1, d2, by = "gene") %>% kable("html") %>% kable_styling()
```


### (4) full_join  

```{r}
full_join(d1, d2, by = "gene") %>% kable("html") %>% kable_styling()
``` 

***

## FILTERING JOIN

### (5) semi_join  

```{r}
semi_join(d1, d2, by = "gene") %>% kable("html") %>% kable_styling()
```

### (6) anti_join    

```{r}
anti_join(d1, d2, by = "gene") %>% kable("html") %>% kable_styling()
```

***

# I have omitted explanations in this cheatsheet on purpose

The reason being that I spent a significant amount of time thinking about sample data.frame d1 and d2 and how to show in a visual snapshot what are the differences between the types of dplyr joining functions. 

#### If you would like more details explanations, please feel free to refer to the long version of this homework 4 [HERE](https://github.com/mylinhthibodeau/STAT545-HW-thibodeau-mylinh/blob/master/stat545-hw4-thibodeau-mylinh/long-version-stat545-hw04-thibodeau-mylinh.md)


