---
title: "hw04-activity5-data-manipulation-cheat-sheet-thibodeau-mylinh"
author: "My Linh Thibodeau"
date: '2017-10-07'
output: github_document
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
knitr::opts_chunk$set(fig.width=12, fig.height=9)
library(knitr)
library(kableExtra)
library(reshape2)
options(knitr.table.format = "html")
```

# Essentials of data manipulations in R

# Matching tasks and functions

| Task          | tidyr function          | reshape2 function   | base R                      |
|---------------|-------------------------|---------------------|-----------------------------|
|               |                         |                     |                             |
|1. Group       | group_by()              | cast()              | aggregate(~ group)          |
|               |                         |                     |                             | 
|2. Split       | spread(), ?separate()   | dcast()             | table()                     |
|               |                         |                     |                             | 
|3. Stack       | gather()                | melt()              |                             | 
|               |                         |                     |                             | 
|4. Join        | _join functions         | reshape::merge_all  | merge(), cbind(), rbind()   |
|               |                         |                     |                             | 
|5. Subset      | filter(), extract()     |                     |                             |  
|               |                         |                     |                             | 
|6. Transpose   | gather()/spread()       |                     | data.frame(t())             |
|               |                         |                     |                             | 
|7. Sort        | arrange()               |                     |                             |
|               |                         |                     |                             |

***

Let's make a small table to illustrate the different funcitons.
```{r}
d1 <- read.table("small_dataframe_example.txt",sep = "\t", header = TRUE)
d1 %>%
  kable("html") %>% kable_styling()
```

## GROUP

Objective: group gene.type and copy.state, when count the number of rows (pairs) and the sum of the copy.number for each pair.

### Tidyr

```{r}
gr_tidy <- select(d1, gene.type, expression, copy.number) %>%
  group_by(gene.type, expression) %>%
  summarize(Nrows = n(), SumCol.copy.number = sum(copy.number))
gr_tidy %>% kable("html") %>% kable_styling()
```

### Reshape

```{r}
gr_resh <-acast(d1, gene.type~expression)
gr_resh %>% kable("html") %>% kable_styling()
```


## SPLIT

Objective: transform the Nrows column (count of gene.type-expression pairs) of gr_tidy so that each group (TS, ONC) is represented by one row, each expression category by one column, and the numbers (Nrows of gr_tidy) represent the number of occurence in each pair (gene.type-expression).

### Tidyr

```{r}
split_tidy <- gr_tidy %>%
  group_by(gene.type, expression) %>%
  select(gene.type, Nrows) %>%
  spread(key=expression, value = Nrows)
split_tidy %>% kable("html") %>% kable_styling()
```


### Reshape

```{r}
split_resh <- gr_tidy %>%
  dcast(gene.type~expression, value.var = "Nrows")
split_resh %>% kable("html") %>% kable_styling()
```

### Base R
 
```{r}
split_R <- with(d1, table(gene.type, expression))
split_R %>% kable("html") %>% kable_styling()
```


## STACK

Objective: display split_resh in a way such that all gene.type - expression possible pairs are displayed in col1 and col2, which col3 returns the count of such pairs (Nrows)

### Tidyr

```{r}
stack_tidy <- split_resh %>%
  gather(key = variable, value = expression, very.low, low,  average, very.high, high)
stack_tidy %>% kable("html") %>% kable_styling()
```


### Reshape

```{r}
stack_resh <- melt(split_resh, id="gene.type") %>%
  arrange(gene.type)
stack_resh %>% kable("html") %>% kable_styling()
```


***

Let's prepare another dataset (d2) for the following exercises.
```{r}
d2 <- read.table("small_dataframe_example_2.txt",sep = "\t", header = TRUE)
d2 %>% kable("html") %>% kable_styling()
```
***

## JOIN

Objective: join d1 and d2 according to the common column gene.

### Tidyr

#### left_join()

```{r}
join_tidy <- left_join(d1, d2, by = "gene")
join_tidy %>% kable("html") %>% kable_styling()
```

### Base R

```{r}
join_R <- merge(d1, d2)
join_R %>% kable("html") %>% kable_styling()
```


## SUBSET

Objective: only take the data for the ONC gene.type.

### Tidyr

```{r}
subset_tidy <- join_tidy %>%
  filter(gene.type == "ONC") 
subset_tidy %>% kable("html") %>% kable_styling()
``` 


## TRANSPOSE

Objective: take the subset_tidy horizontal table and transpose it to a vertical position such that the rows become columns and vice versa.

### Base R

```{r}
transpose_R <- data.frame(t(subset_tidy))
transpose_R %>% kable("html") %>% kable_styling()
```

I note that the columns are called X1 and X2 instead of ONC and ONC. Oh well, we get the general idea anyway !

## SORT

Objective: sort the table according to ascending copy.number

### Tidyr

```{r}
join_tidy %>%
  arrange(copy.number) %>% kable("html") %>% kable_styling()
```


***                
REFERENCES
Reshape2 manual [here](https://www.rdocumentation.org/packages/reshape2/versions/1.4.2)
Reshape2 intro [here](http://seananderson.ca/2013/10/19/reshape.html)
Reshape reference [here](http://had.co.nz/reshape/), [here](https://stackoverflow.com/questions/8091303/simultaneously-merge-multiple-data-frames-in-a-list)
Blog on data manipulation [here on r-statistics](https://www.r-statistics.com/2012/01/aggregation-and-restructuring-data-from-r-in-action/), and [here on Oregon University](http://library.open.oregonstate.edu/computationalbiology/chapter/reshaping-and-joining-data-frames/).
Panda can also offer options, as exemplified [here](https://pandas.pydata.org/pandas-docs/stable/merging.html), but I ran out of time to look into this further.
