---
title: "stat545-hw04-thibodeau-mylinh"
author: "My Linh Thibodeau"
date: '2017-10-07'
output: github_document
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
knitr::opts_chunk$set(fig.width=12, fig.height=9)
library(knitr)
library(kableExtra)
options(knitr.table.format = "html")
#install.packages("scales")
library(scales)
#install.packages("tidygenomics")
library(tidygenomics)
```

# Genomic dataset - A few clarifications

Vincenzo Coia has approved my request to use published genomic data for the next assignments, therefore, I want to provide a few clarifications:

* I have tried to introduce some basic explanations about the genomic dataset, but obviously, this is not a genetics course and my objective is to explore and learn how to use R and its packages, not to teach complex notions of cancer genomic analysis. Therefore, I don't expect people to understand what the data and plots represent if they haven't studied in related fields.  
* I would recommend you make abstraction of the underlying biological context and simply try to read this homework based on the variable types (e.g. copy.category is a categorical variable like "gain" or "loss") rather than what they represent.

## SOURCE OF DATA: The Cancer Genome Atlas (TCGA)

After trying to complete this homework with the genomic data previously used in [homework 3](https://github.com/mylinhthibodeau/STAT545-HW-thibodeau-mylinh/tree/master/stat545-hw3-thibodeau-mylinh), I realized that it was impossible for me to do some of the exercises because I didn't have the right "data.frame" format for it. 

# Setting up the data

## Getting the data ready

```{r}
brca_rna <- read.table("/Users/mylinh/Documents/published_data/tcga/brca/tcga/data_RNA_Seq_v2_expression_median.txt", header = TRUE, nrows =50)[ ,c("Hugo_Symbol", "TCGA.3C.AAAU.01", "TCGA.3C.AALI.01", "TCGA.3C.AALJ.01")]
brca_rna
```

