---
title: "gapminder-exploration-phase2"
author: "My Linh Thibodeau"
date: '2017-09-22'
output: html_document
  toc: true
---

```{r}
suppressPackageStartupMessages(library(tidyverse))  # The tidyverse contains ggplot2!
suppressPackageStartupMessages(library(gapminder))
knitr::opts_chunk$set(fig.width=4, fig.height=3)
```

### Smell test of data
Here are some ways to explore some characteristics of the gapminder dataset.

Using the 
```{r}
class(gapminder)
typeof(gapminder)
str(gapminder)
```

I found the Help documentation very useful to understand better the concept of an object in R coding.
```{r}
?class
?tbl_df
?tbl
?data.frame
?typeof
?str
```

Therefore, I now understand that gapminder is a tibble. It seems like using tibbles over classic data.frame() has several advantages, like avoiding problems such as accidental conversion of string as factor. I read this interesting [wiki page](http://www.sthda.com/english/wiki/tibble-data-format-in-r-best-and-modern-way-to-work-with-your-data) on the topic.

The gapminder data set has 3 classes  
* tbl_df  
* tbl  
* data.frame  

**I found class() to be the most informative function for me**   

The class function told me that there were 6 variables (columns) and 1704 observations (rows) in gapminder dataset, but I could also use this:
```{r}
ncol(gapminder)
nrow(gapminder)
```

It also revealed that the specific data type of each variable!  
* country  : Factor  
* continent: Factor  
* year     : int  
* lifeExp  : num  
* pop      : int  
* gdpPercap: num  

Here are some other useful functions to learn more about the dataset. I will not explain what each simple function does at this time, but please be reassured that I used the Help function to understand the use and output of these functions. If you have a specific question about the dataset, you can use more specific functions as well, like dim() which provides the dimension of the data tibble.
```{r}
dim(gapminder)
```

### Explore individual variables
> To ensure I address all the requirements of this assignment, I have decided to include the STAT545 homework 2 questions above my answers in bold. 

*Q? What are possible values (or range, whichever is appropriate) of each variable?*  
I have used the summary() function to know the possible values and range of each variable.
```{r}
View(gapminder)
summary(gapminder)
```

Another way is to use individual functions to obtain the information. Below, these functions respectively informs on the minimum value, maximum value, range of value (minimum and maximum) and all possible distinct values of specific columns. 
```{r}
min(gapminder$lifeExp)
max(gapminder$lifeExp)
range(gapminder$gdpPercap)
distinct(gapminder, country)
```

I also found this ![website](https://info201-s17.github.io/book/introduction-to-data-frames.html) useful. 

*Q? What values are typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand.Feel free to use summary stats, tables, figures. We’re NOT expecting high production value (yet).*  

If we group data according to their column, another useful function is summarise() to get some statistics on a variable, but data needs to be grouped before doing so. For example, if we group by continent, we can use the function to know the mean year of entry, lifeExp, pop and gdpPercap for each continent.
```{r}
gapminder %>%
  group_by(continent) %>%
  summarise_if(is.numeric, mean, na.rm=TRUE)
```
Again, I find that the most useful function is summary(), as it provides the range (minimum, maximum), and summary statistics such as 1st quartile, mean and 3rd quartile.
```{r}
View(gapminder)
summary(gapminder)
```

```{r}
gapminder %>% 
  group_by(country) %>% 
  summarise_if(is.numeric, median)
```


### Explore various plot types

