---
title: "gapminder-exploration-phase2"
author: "My Linh Thibodeau"
date: '2017-09-22'
output: html_document
  toc: true
---

```{r}
suppressPackageStartupMessages(library(tidyverse))  # The tidyverse contains ggplot2!
suppressPackageStartupMessages(library(gapminder))
knitr::opts_chunk$set(fig.width=4, fig.height=3)
```

### Smell test of data
Here are some ways to explore some characteristics of the gapminder dataset.

Using the 
```{r}
class(gapminder)
typeof(gapminder)
str(gapminder)
```

I found the Help documentation very useful to understand better the concept of an object in R coding.
```{r}
?class
?tbl_df
?tbl
?data.frame
?typeof
?str
```

Therefore, I now understand that gapminder is a tibble. It seems like using tibbles over classic data.frame() has several advantages, like avoiding problems such as accidental conversion of string as factor. I read this interesting [wiki page](http://www.sthda.com/english/wiki/tibble-data-format-in-r-best-and-modern-way-to-work-with-your-data) on the topic.

The gapminder data set has 3 classes  
* tbl_df  
* tbl  
* data

```{r}

```

