---
title: "stat547-hw06-thibodeau-mylinh"
author: "My Linh Thibodeau"
date: '2017-11-05'
output: html_document:
  keep_md: true
---


```{r message=FALSE}
#suppressPackageStartupMessages(library(plyr))
suppressPackageStartupMessages(library(tidyverse))
suppressWarnings(library(tidyverse))
knitr::opts_chunk$set(fig.width=13, fig.height=8)
suppressWarnings(suppressMessages(library(knitr)))
suppressWarnings(suppressMessages(library(kableExtra)))
suppressWarnings(options(knitr.table.format = "markdown"))
library(readr)
suppressWarnings(suppressMessages(library(forcats)))
library(purrr)
library(repurrrsive)
library(tibble)
library(devtools)
library(singer)
library(listviewer)
library(jsonlite)
library(dplyr)
library(tibble)
library(rlang)
#install.packages("ggmap")
library(ggmap)
#install.packages("ontologyIndex")
library(ontologyIndex)
#install.packages("ontologySimilarity")
library(ontologySimilarity)
#install.packages("ontologyPlot")
library(ontologyPlot)
```

```{r}
getNamespaceExports("ontologyIndex") 
getNamespaceExports("ontologySimilarity") 
```


First step, get the Human Disease Ontology as explained [here](https://github.com/DiseaseOntology/HumanDiseaseOntology/blob/master/src/ontology/README-editors.md), I had to clone the repository (which I called HumanDiseaseOntology_git, as suggested by the github README document).
```{r}
HumanDO <- get_ontology("HumanDiseaseOntology_git/src/ontology/HumanDO.obo")
View(HumanDO)
```

Second, get the Human Phenotype Ontology information:
```{r}
data(hpo)
glimpse(hpo) 
View(hpo)
data(go)
View(go)
```

There is a specific file with information on the Online Mendelian Inheritance in Man (OMIM) annotations that I would like to look into with the ontologyIndex package, but the formal of the file is OWL, and I need to convert it to OBO first. I have used this github repository [here](https://github.com/owlcollab/oboformat) to find information about ROBOT and how to convert the file. 

```{r}
doid_plus_omim <- get_ontology("HumanDiseaseOntology_git/src/ontology/DO_NON_Production_Files/doid-plus-omim.obo")
```


# Homework instructions

**Pick (at least) two of the six (numbered) topics below and do one of the exercise prompts listed, or something comparable using your dataset of choice.**



```{r}

```

